---
title: "Supervised Multiblock analyses"
author: "Florence Mehl"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  html_document: 
    toc: TRUE
    theme: paper
    higlight: tango
    code_folding: "hide"

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(MBAnalysis)
library(CCA)
library(ggplot2)
library(ggrepel)
library(reshape2)
library(mixOmics)
library(RVAideMemoire)
library(abind)
library(ConsensusOPLS)
library(BiocParallel)

```

# Context and dataset introduction

This dataset originates from a nutrigenomic study in mice (Martin et al., 2007), designed to investigate how dietary fatty acid composition influences liver lipid profiles and hepatic gene expression.

**Experimental Design**  

Forty mice were cross-classified according to two experimental factors, each with multiple levels:

- Genotype (2 levels):  
Wild-type (WT)  
PPARα knockout (PPARα -/-)  

- Diet (5 levels):

REF: Reference diet (50/50 corn and colza oils)  
COC: Saturated fatty acid-rich diet (hydrogenated coconut oil)  
SUN: Omega-6-rich diet (sunflower oil)  
LIN: Omega-3-rich diet (linseed oil)  
FISH: Mixed diet (corn/colza/enriched fish oils in 43/43/14 ratio)  

Each genotype-diet combination includes four biological replicates.

**Variables Collected**  

Two sets of variables were measured for each mouse:

- Gene Expression:  

Expression levels of 120 selected hepatic genes, chosen from ~30,000 candidates for their relevance to nutritional regulation.  
Measured using nylon macroarrays with radioactive labeling.  

- Lipid Composition:  

Relative concentrations (percentages) of 21 hepatic fatty acids.  
Quantified via gas chromatography.  


```{r import and format dataset, message=FALSE, warning=FALSE}

data("nutrimouse")
genes <- nutrimouse$gene
lipids <- nutrimouse$lipid
metadata <- data.frame(genotype = nutrimouse$genotype, diet = nutrimouse$diet)
metadata$sample_name <- paste0(rownames(metadata), "_", metadata$genotype, "_", metadata$diet)
rownames(genes) <- metadata$sample_name
rownames(lipids) <- metadata$sample_name

```

# Discriminant analysis of genotypes

## Block PLS-DA Analysis

- Format lipid and gene data as a list of dataframes.
- Define the outcome variable (genotype) as a factor.
- Run block.plsda() from the mixOmics package to model genotype discrimination.  

```{r data preparation, message=FALSE, warning=FALSE}

# prepare data
blockPLS_data <- list(genes=genes, lipids=lipids)
genotype <- as.factor(metadata$genotype)

# run analysis
blockPLS_res <- block.plsda(X = blockPLS_data, Y = genotype, design = "full", all.outputs = T, ncomp = 10)

```

## Latent Variable Selection

- Use perf() to evaluate model performance across components and visualise.
- Re-run block.plsda() using the optimal number of latent variables.

```{r choice the size of model, message=FALSE, warning=FALSE, fig.align='center'}

bp <- MulticoreParam(workers = 4)

blockPLS_perf <- perf(blockPLS_res, validation = 'Mfold', folds = 7, nrepeat = 1, auc = TRUE, BPPARAM = bp, progressBar = FALSE)

blockPLS_perf <- perf(
  blockPLS_res,
  validation = "Mfold",
  folds = 7,
  nrepeat = 100,
  auc = TRUE,
  BPPARAM = bp,
  seed = 1
)

plot(blockPLS_perf)

blockPLS_res <- block.plsda(X = blockPLS_data, Y = genotype, design = "full", all.outputs = T, ncomp = 2)

```

## Model Significance

- Perform a cross-validation to estimate how well your fitted MBPLSDA/DIABLO model generalises to new data using DIABLO.cv() from the RVAideMemoire package.  
- Perform a permutation test to test whether your observed model performance is significantly better than chance using DIABLO.test() from the RVAideMemoire package.

```{r permutation test, message=FALSE, warning=FALSE}

# Cross-validation with DIABLO.cv()
blockPLS_cv <- DIABLO.cv(blockPLS_res,
                         validation = c("Mfold"),
                         k = 7,
                         repet = 100)

blockPLS_cv

# Permutation test with DIABLO.test()
blockPLS_permtest <- DIABLO.test(blockPLS_res,
                                 nperm = 100,
                                 progress = TRUE)

blockPLS_permtest

```

## Explained Variance?

Plot explained variance per block and globally across latent variables.
- Percentage of variance explained in each block by each LV: AVE_X
- How strongly each block contributes to global LV structure: AVE[["AVE_outer"]]

```{r explained variance, message=FALSE, warning=FALSE, fig.align='center'}

blockPLS_expl <- do.call("rbind",blockPLS_res$AVE$AVE_X[1:2])

# extract the cumulative global explained variance
ave_outer_cum <- blockPLS_res$AVE[["AVE_outer"]]
# convert to incremental (per component)
ave_outer_inc <- data.frame(comp1 = ave_outer_cum[1], comp2= diff(ave_outer_cum), row.names = "global")

blockPLS_expl <- rbind(blockPLS_expl, ave_outer_inc)


```

## Sample Space Visualization

- Use plotIndiv(..., block = "weighted.average") to visualize sample distributions in latent variable space.

```{r comdim all samples scores, message=FALSE, warning=FALSE, fig.align='center'}

plotIndiv()

```

## Variable Contributions

- Use plotVar() to identify discriminant genes and lipids contributing to genotype separation.  

```{r comdim all samples loadings, message=FALSE, warning=FALSE, fig.align='center'}

plotVar()

```

# Consensus OPLS Discriminant Analysis of genotypes

## Data Preparation

- Organize lipid and gene datasets as a list of dataframes (one per block).  
- Encode the outcome variable (genotype).   

```{r data prep consensusOPLS-DA, message=FALSE, warning=FALSE, fig.align='center'}

COPLS_data <- list(genes=as.matrix(genes), lipids=as.matrix(lipids))
COPLS_data <- lapply(COPLS_data, scale)
genotype <- metadata$genotype

```

## Consensus OPLS Analysis

- Fit the Consensus OPLS model using ConsensusOPLS() from the ConsensusOPLS package.  
- Model the relationship between the predictor blocks (lipids, genes) and the genotype outcome.  

```{r consensusOPLS-DA, message=FALSE, warning=FALSE, fig.align='center'}
COPLS_res <- ConsensusOPLS(
  data = COPLS_data,
  Y = genotype,
  maxPcomp = 1,
  maxOcomp = 1,
  modelType = "da",
  cvType = "nfold",
  nfold = 40,
  nperm = 100,
  verbose = T,
  kernelParams = list(type='p', params = c(order=1.0))
)
```


## Model significance

- Assess statistical significance using permutation tests and visualize with histograms of
Q², dQ², and R²Y values from permuted models.  
- The results of permutations can be found in `COPLS_res$permStats`.
- The results for the optimal model can be found in `COPLS_res$optimal$modelCV` and `COPLS_res$optimal$modelCV$cv`
- plot Q2 permutations
- plot DQ2 permutations
- plot R2Y permutations

```{r permutations, message=FALSE, warning=FALSE, fig.align='center'}

plotQ2()
plotDQ2()
plotR2()

```

## Block Contributions

-Plot block contributions as a bar plot to see which omics layer drives genotype discrimination.  

```{r consensus OPLS-DA contributions, message=FALSE, warning=FALSE, fig.align='center'}

plotContribution()

```

## Sample Space Visualization

- Plot scores to visualize sample distribution in the space of predictive and orthogonal latent variables.  

```{r consensus OPLS-DA scores, message=FALSE, warning=FALSE, fig.align='center'}

plotScores()

```

## Variable Contributions

- Plot loadings: predictive vs orthogonal components
- Interpret biological relevance

```{r consensus OPLS-DA loadings, message=FALSE, warning=FALSE, fig.align='center'}

plotLoadings()

```

- Identify most discriminant genes plotting VIP vs loadings on predictive component

```{r consensus OPLS-DA VIP loadings, message=FALSE, warning=FALSE, fig.align='center'}

plotVIP()

```

## Predicting with consensusOPLS

Train a Consensus OPLS-DA model to discriminate between WT and PPAR samples using a training set and evaluate its performance on an independent test set.

### Data Splitting

- Select 30 samples for training and 10 samples for testing.  
- Organize lipid and gene datasets as lists of matrices for training and testing.  
- Scale each block separately for training and test sets.  
- Define the outcome variable (genotype) for the training set.  

```{r consensus OPLS-DA external validation data preparation, message=FALSE, warning=FALSE, fig.align='center'}

```

### Consensus OPLS Analysis

- Fit the Consensus OPLS model on the training set.  

```{r consensus OPLS-DA external validation model, message=FALSE, warning=FALSE, fig.align='center'}

```

### Model Evaluation

- Assess statistical significance using permutation tests.  
- visualize plotting Q², dQ², and R²Y values from permuted models.  
- Observe block contributions, scores and loadings.  

```{r consensus OPLS-DA external validation evaluation, message=FALSE, warning=FALSE, fig.align='center'}

```

### External Validation on Test Set

- Predict class labels for the independent test samples using predict().  
- Compare predicted classes and predicted Y values against the true genotypes.  
- Summarize in a table for accuracy evaluation.  

```{r consensus OPLS-DA external validation on test set, message=FALSE, warning=FALSE, fig.align='center'}

test.COPLS_res <- predict(object = train.COPLS_res,
                          newdata = test.data)

# results

data.frame(test.COPLS_res$class,
           Y.pred = test.COPLS_res$Y,
           True.genotype=metadata$genotype[test.observations])

```




